---
title: "Accesing Single-cell API REST from R"
author: 
- Alicia Gómez
- Eduardo Salmerón
date: "`r format(Sys.time(), '%d/%m/%y')`"
output: 
  html_document:
    theme: spacelab
    highlight: kate
    df_print: paged
    toc: true
    toc_float: true
    number_sections: true
---

<style>
body {
text-align: justify}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
source("/home/aligo/API/functions.R")
source("/home/aligo/API/scGCNsFunctions.R")
APIpath <- "http://194.4.102.168:5000/"
library(gridExtra)
# https://medium.com/@traffordDataLab/querying-apis-in-r-39029b73d5f1
```

# Requirements

```{r}
library(tidyverse)
library(httr)
library(jsonlite)
library(data.table)
library(RCurl)
library(rjson)
```


# Get a list of possible values for a metadata

## Disease

```{r}
request <- GET(url = paste0(APIpath, "project/metadata/disease"))
response <- content(request, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(response, flatten = TRUE)
df$term[1:50]
```

## cell_type

```{r}
request <- GET(url = paste0(APIpath, "project/metadata/cell_type"))
response <- content(request, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(response, flatten = TRUE)
df$term[1:50]
```


# Get percentiles of the projects given a filter

```{r, eval=FALSE}
request <- GET(url = paste0(APIpath, "percentiles?filters=%7B%0A%20%20%22cell_types%22%3A%20%5B%0A%20%20%20%20%22memory%20B%20cell%22%2C%0A%20%20%20%20%22blood%20cell%22%0A%20%20%5D%2C%0A%20%20%22gen_names%22%3A%20%5B%0A%20%20%20%20%22ENSG00000287846%22%2C%0A%20%20%20%20%22ENSDARG00000034326%22%0A%20%20%5D%0A%7D"))
response <- content(request, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(response) %>% 
  data.frame()
df
```



# Get a list of projects

## List all projects

```{r}
request <- GET(url = paste0(APIpath, "project?"))
response <- content(request, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(response)
df <- as.data.frame(do.call(cbind, lapply(df, function(x) unlist(x))))
dim(df)
df[1:5, 1:5]
```


## List projects given a filter

```{r}
request <- GET(url = paste0(APIpath, "project?cell_type=DopaminergicNeuron"))
response <- content(request, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(response)
df <- as.data.frame(do.call(cbind, lapply(df, function(x) unlist(x))))
df
```


# Get downloads

```{r}
request <- GET(url = paste0(APIpath, "project/downloads/E-MTAB-7303"))
response <- content(request, as = "text", encoding = "UTF-8")
df <- rjson::fromJSON(response, simplify=T)
df <- setDT(as.data.frame(do.call(cbind, lapply(df, function(x) unlist(x)))), keep.rownames=T)
colnames(df) <- c("name", "link")
df$link <- as.character(df$link)
df
```

```{r}
metadata <- getURL(df$link[df$name=="experimentDesignLink"])
metadata <- read.csv(textConnection(metadata), sep="\t")
metadata[1:5, 1:5]
```


# Implementation

## Step 1: obtaining the files for 15 different projects

```{r, eval=FALSE}
source("/home/aligo/API/functions.R")
myPath <- "/home/aligo/API/"
APIpath <- NULL
# sceaProjects <- c("E-MTAB-7303", "E-CURD-11", "E-CURD-12", "E-CURD-13", "E-ENAD-20", "E-GEOD-110499", "E-GEOD-149689", "E-MTAB-5802", "E-MTAB-7381", "E-MTAB-8559")
# hcaProjects <- c("8c3c290d-dfff-4553-8868-54ce45f4ba7f", "116965f3-f094-4769-9d28-ae675c1b569c", "f8aa201c-4ff1-45a4-890e-840d63459ca2", "4d6f6c96-2a83-43d8-8fe1-0f53bffd4674")

ID <- "E-HCAD-13"

for (ID in hcaProjects) {
  print(ID)
  
  # Create dir if doesn't exist
  projectName <- ID
  workingPath <- paste0(myPath, "/", projectName)
  
  if(!(dir.exists(workingPath))) {
    cat("Creating", workingPath, "directory\n")
    dir.create(workingPath)
  }
  
  setwd(workingPath)
  myFiles <- list.files(workingPath)
  index <- grep(".rds", myFiles)

  if (length(index)==1) {
    cat("We already have the matrix created\n")
    matrix <- readRDS(paste0(workingPath, "/", myFiles[index]))
    cat("Creating metadata file\n")
    metadata <- getMetadata(df, matrix, workingPath)
    
  } else if (length(index)==0) {
    cat("Creating matrix file\n")
    matrix <- getMatrix(ID, workingPath, APIpath)
    cat("Creating metadata file\n")
    metadata <- getMetadata(ID, workingPath, APIpath)
  }
  
}

```


```{r}
matrix <- readRDS("~/API/E-MTAB-7303/E-MTAB-7303.expression_tpm.rds")
dim(as.matrix(matrix))
matrix[1:5, 1:5]
```


```{r}
metadata <- readRDS("~/API/E-MTAB-7303/ExpDesign-E-MTAB-7303.rds")
dim(metadata)
metadata[1:5, ]
```


## Step 2: classify biological and technical covariates

```{r}
json_data <- rjson::fromJSON(file="/home/aligo/API/posible_values.json")
sampleFeatures <- names(json_data[[1]])
expFeatures <- names(json_data[[2]])
sort(sampleFeatures)
```


```{r}
ctID <- "Assay"
sampleID <- "individual"
ctCov <- c("cell.type") # inferred cell type?
subCov <- c("cluster")
samplingCovs <- c("growth.condition", "metastatic.site", "organism", "organism.part", "sampling.site", "stimulus")
bioCovs <- c("age", "ancestry.category", "cell.cycle.phase", "cell.line", "developmental.stage", "disease", "disease.staging", "ethnic.group", "phenotype", "post.conception.age", "sex")
technicalCovs <- c("block", "microwell.plate", "organism.status", "passage", "plate", "protocol", "sampling.time.point", "time", "well", "well.information")
myList <- list(ctID, sampleID, ctCov, subCov, samplingCovs, bioCovs, technicalCovs)
# saveRDS(myList, "/home/aligo/API/myList.rds")
```



## Estimate the number of possible combinations for each project

```{r}
samplingCovs <- c("organism", "organism.part", "sampling.site", "biopsy.site", "metastatic.site", "developmental.stage", "cell.line", "cell.type", "inferred.cell.type - ontology.labels")
myPath <- "/home/aligo/API/"
projects <- list.dirs(myPath, recursive=FALSE)
names <- gsub(myPath, "", projects)
names <- gsub("/", "", names)
metadataList <- list()
results <- list()

for (i in 1:length(projects)) {
  
 cat("- Project_ID:", names[i], "\n")
  repo <- grep("E", names[i])
  
  if(length(repo)>0) {
    cat("- Repository: SCEA \n")
  } else {
    cat("- Repository: HCA \n")
  }
    
  metadata <- list.files(projects[i], pattern="ExpDesign", full.names=T)
  metadata <- readRDS(metadata)
  metadataList[[i]] <- metadata
  
  ## Remove cols with the same value for all cells
  colsToRemove <- c()
   
  for (j in 1:ncol(metadata)) {
    term <- as.character(metadata[, j])
     counts <- table(term)
   
     if (length(counts)==1) {
       colsToRemove <- c(colsToRemove, j)
     }
   }
   
  metadata <- metadata[, -colsToRemove]
  
  ## Looking for sampling covs
  index <- which(colnames(metadata) %in% samplingCovs)
   
  if(length(index)>0) {
    ## We split our dataframe based on sampling covs
    cols <- list()
    
    for (k in 1:length(index)) {
      cols[[k]] <- metadata[, colnames(metadata)[index[k]]]
    }
    
    mySplit <- split(metadata, cols, drop=TRUE)
    
    ## Show results
    cat("- Sampling covs:", paste0(colnames(metadata)[index], collapse=", "), "\n")
    combs <- lapply(mySplit, function(x) nrow(x))
    cat("- Number of cells by combination:", paste0(names(combs), ", " , combs, " cells; "))
    cat("\n\n")
        
  } else {  
    ## Show results
    if(!is.null(dim(metadata))) {
      cat("- Number of cells:", nrow(metadata), "\n\n")
    } else {
       cat("- Number of cells:", length(metadata), "\n\n")
    }
    
  }
}

```


## My first GCN experiment

```{r}
myResults <- as.data.frame(matrix(numeric(), nrow = 0, ncol = 5))
projects <- c("E-MTAB-7303", "E-ENAD-20", "f8aa201c-4ff1-45a4-890e-840d63459ca2")

for (project in projects) {
  dirs <- list.dirs(paste0("/home/aligo/API/", project), recursive=FALSE)
  
  if(length(dirs)==1) {
    dirs <- list.dirs(paste0("/home/aligo/API/", project))[-1]
    dirs <- dirs[-grep("Plots", dirs)]
    dirs <- dirs[-grep("Net", dirs)]
    dirs <- dirs[-grep("Residuals", dirs)]
    dirs <- dirs[grep("iter", dirs)]
    
  }

  for (dir in dirs) {
    name <- gsub(paste0("/home/aligo/API/", project, "/"), "", dir)
    sum <- mySummary(name, paste0(dir, "/"))
    myResults <- rbind(myResults, sum)
  }
}

rownames(myResults) <- c()
myResults
```




